Parametricity for Primitive Nested Types and GADTs

### 2020-12-13 free-theorems-nesteds/20-mscs.tex
add biblio and new mu rule

\begin{abstract}
This paper considers parametricity and its consequent free theorems
for nested data types. Rather than representing nested types via their
Church encodings in a higher-kinded or dependently typed extension of
System F, we adopt a functional programming perspective and design a
Hindley-Milner-style calculus with primitives for constructing nested
types directly as fixpoints. Our calculus can express all nested types
appearing in the literature, including truly nested types. At the
level of terms, it supports primitive pattern matching, map functions,
and fold combinators for nested types. Our main contribution is the
construction of a parametric model for our calculus. This is both
delicate and challenging. In particular, to ensure the existence of
semantic fixpoints interpreting nested types, and thus to establish a
suitable Identity Extension Lemma for our calculus, our type system
must explicitly track functoriality of types, and cocontinuity
conditions on the functors interpreting them must be appropriately
threaded throughout the model construction. We also prove that our
model satisfies an appropriate Abstraction Theorem, as well as that it
verifies all standard consequences of parametricity in the presence of
primitive nested types. In particular, we give several examples
illustrating how how our model can be used to derive useful free
theorems for programs over nested types. {\color{red} Add ainfo about
  GADTs and implementation}
\end{abstract}

@@

  Send to LMCS? Editor?

Can bind variables of arity > 0 in Nat-types? Improves expressibilty?
Worth it?}

@@

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;\ol{\gamma^0},\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash H}$}
\TIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F[\ol{\gamma := H}])\,\ol{G}$}  
\DisplayProof
\end{array}\]

@@

Definition~\ref{def:wftypes} allows the formation of all of the
(closed) nested types from the introduction:
\[\begin{array}{lll}
\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
\beta)\,\alpha\\ 
\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
\phi\,(\beta \times \beta))\,\alpha\\
\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
\times \phi\,(\phi\beta))\,\alpha 
\end{array}\]
Each of these types can be considered either functorial in $\alpha$ or
not, according to whether $\alpha$ is in the functorial or
non-functorial context in which it is formed. In particular, if
$\emptyset;\alpha \vdash \mathit{List} \,\alpha$, then $\vdash
\Nat^\alpha \onet\, (\mathit{List}\,\alpha)$ is well-formed; if
$\alpha;\emptyset \vdash \List \,\alpha$, then it is not.  Similarly,
if $\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta
\times \gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is
well-formed, but $\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\,
\alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is not. Although types
can be functorial in variables of arity greater than $0$, the body $F$
of a type $(\mu \phi.\lambda \ol\alpha.F)\ol G$ cannot. For example,
the type $\mathit{GRose}\, \phi \, \alpha = \mu\beta . \alpha \times
\phi \beta$ can be functorial or not in $\alpha$, but cannot be
functorial in $\phi$. Moreover, $\mathit{GRose}\,\phi\,\alpha$ cannot
be the (co)domain of a $\Nat$-type representing a natural
transformation functorial in $\phi$, and the type $\mu \phi. \lambda
\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
system. These restrictions ensure that the set and relational
interpretations of types in Section~\ref{sec:type-interp} are
functors. In Section~\ref{sec:conclusion} we discuss how
more expressive types might be accommodated.

Definition~\ref{def:wftypes} allows well-formed types to be 
functorial in no variables
Functorial variables
%in a well-formed type $\tau$
can also be demoted to non-functorial status: if\,$F[\phi :== \psi]$
is the textual replacement of $\phi$ in $F$, then $\Gamma, \psi^k;
\Phi \vdash F[\phi^k :== \psi^k]$ is derivable whenever $\Gamma; \Phi,
\phi^k \vdash F$ is. The proof is by induction on the structure of
$F$. In addition to textual replacement, we also have a
substitution operation on types.

@@

{\color{red} Check no change needed for new $\mu$-rule in next def.}
\begin{defi}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
\vdash F$ with $|\ol\alpha| = k$, then the operation $H[\phi :=_{\ol
    \alpha} F]$ of {\em second-order type substitution along
  $\ol\alpha$} is defined by induction on $H$ as follows:
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, (G[\phi :=_{\ol{\alpha}} F]) \,(K[\phi
  :=_{\ol{\alpha}} F])\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{defi}
\noindent
It is not hard to see that $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
  F]$. In addition, if \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma;
\ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap
\ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash H[\phi
  :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.

@@

       \[\begin{array}{c}
       \AXC{$\Gamma; \ol{\alpha} \vdash F$}
       \AXC{$\Gamma; \ol{\alpha} \vdash G$}
       \AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
       \TIC{$\Gamma; \emptyset
         \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
       \DisplayProof\vspace*{-0.05in}
       \\\\
       \AXC{$\ol{\Gamma;\Phi \vdash K}$}
       \AXC{$\Gamma; \emptyset
         \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
       \TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
         G[\overline{\alpha := K}]$}
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash
           G}$}
       \TIC{$\Gamma; \emptyset
         ~|~\emptyset
         \vdash \map^{\ol{F},\ol{G}}_H :
         \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
         (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
             :=_{\ol{\beta}} G}])$} 
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi, \ol{\alpha},\ol{\gamma} \vdash H$}
       \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
         \Nat^{\ol{\beta},\ol{\gamma}} H[\phi :=_{\ol{\beta}} (\mu
           \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
         \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi,\ol{\alpha}, \ol{\gamma} \vdash H$}
       \AXC{$\Gamma; \ol{\beta}, \ol{\gamma} \vdash F$}
       \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
         \Nat^\emptyset\; (\Nat^{\ol{\beta}, \ol{\gamma}}\,H[\phi
           :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
           \ol{\gamma}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \end{array}\]

@@

       \caption{Well-formed terms {\color{red} Change last three
           judgements for new $\mu$ rule???}}\label{fig:terms} \vspace*{-0.00in}
\end{adjustbox}
%       \vspace*{-0.25in}
\end{figure*}

@@

{\color{red} Change the rest of the section to agree with new term
  formation rules.}  The ``extra'' functorial variables $\ol{\gamma}$
in the rules for $\map^{\ol{F},\ol{G}}_H$, $\tin_H$, and $\fold^F_H$
(i.e., those variables not affected by the substitution of $\phi$)
deserve comment. They allow us to map or fold polymorphic functions
over nested types. Suppose, for example, we want to map the
polymorphic function $\mathit{flatten} : \Nat^\beta
(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ over a list.  Even in
the absence of extra variables the instance of $\map$ required to map
each non-functorial monomorphic instantiation of $\mathit{flatten}$
over a list of perfect trees is well-typed:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
\end{array}\]
But in the absence of $\ol \gamma$, the instance
\[\Gamma;\emptyset~|~\emptyset \vdash
\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
\beta)))\] required to map the {\em polymorphic} $\mathit{flatten}$
function over a list of perfect trees is not: indeed, the functorial
contexts for $F$ and $G$ in the rule for $\map^{F,G}_H$ would have to
be empty, but because the polymorphic $\mathit{flatten}$ function is
natural in some variable, say $\delta$, it cannot possibly have a type
of the form $\Nat^\emptyset F\, G$ as would be required for it to be
the function input to $\map$. Untypeability of this instance of $\map$
is unsatisfactory in a polymorphic calculus, where we naturally expect
to be able to manipulate entire polymorphic functions rather than just
their monomorphic instances, but the ``extra'' variables $\ol \gamma$
remedy the situation, ensuring that the instance of $\map$ needed to
map the polymorphic $\mathit{flatten}$ function is typeable as
follows:

@@

@@

Unfortunately, our calculus cannot express types of recursive
functions --- such as a concatenation function for perfect trees or a
zip function for bushes --- that take as inputs a nested type and an
argument of another type, both of which are parameterized over the
same variable. The fundamental issue is that recursion is expressible
only via $\fold$, which produces natural transformations in some
variables $\ol\alpha$ from $\mu$-types to other functors $F$. The
restrictions on $\Nat$-types entail that $F$ cannot itself be a
$\Nat$-type containing $\ol{\alpha}$, so, e.g., $\Nat^\alpha
(\mathit{PTree}\,\alpha) \,(\Nat^\emptyset (\mathit{PTree}\,\alpha)\,
(\mathit{PTree}\,(\alpha \times \alpha)))$ is not well-typed.
Uncurrying gives $\Nat^\alpha (\mathit{PTree}\,\alpha \times
\mathit{PTree}\,\alpha)\, (\mathit{PTree}\,(\alpha \times \alpha))$,
which is well-typed, but $\mathsf{fold}$ cannot produce a term of this
type because $\mathit{PTree}\,\alpha \times \mathit{PTree}\,\alpha$ is
not a $\mu$-type.  Our calculus can, however, express types of
recursive functions that take multiple nested types as arguments,
provided they are parameterized over disjoint sets of type variables
and the return type of the function is parameterized over only the
variables occurring in the type of its final argument.  Even for ADTs
there is a difference between which folds over them we can type when
they are viewed as ADTs (i.e., as fixpoints of first-order functors)
versus as proper nested types (i.e., as fixpoints of higher-order
functors). This is because, in the return type of $\mathsf{fold}$, the
arguments of the $\mu$-type must be variables bound by $\Nat$.  For
ADTs, the $\mu$-type takes no arguments, making it possible to write
recursive functions, such as a concatenation function for lists of
type $\alpha ; \emptyset \vdash \Nat^\emptyset \ (\mu \beta. \onet +
\alpha \times \beta)\, (\Nat^\emptyset (\mu \beta. \onet + \alpha
\times \beta) \, (\mu \beta. \onet + \alpha \times \beta))$.  This is
not possible for nested types --- even when they are semantically
equivalent to ADTs.

@@

\item If \,$\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda

@@

  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu  \phi.\lambda
    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi,\ol\gamma 

@@

  \setsem{\Gamma;\Phi,\ol\gamma \vdash (\mu

@@

  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho} \to (\mu

>>>>

\begin{abstract}
This paper considers parametricity and its consequent free theorems
for nested data types. Rather than representing nested types via their
Church encodings in a higher-kinded or dependently typed extension of
System F, we adopt a functional programming perspective and design a
Hindley-Milner-style calculus with primitives for constructing nested
types directly as fixpoints. Our calculus can express all nested types
appearing in the literature, including truly nested types. At the
level of terms, it supports primitive pattern matching, map functions,
and fold combinators for nested types. Our main contribution is the
construction of a parametric model for our calculus. This is both
delicate and challenging. In particular, to ensure the existence of
semantic fixpoints interpreting nested types, and thus to establish a
suitable Identity Extension Lemma for our calculus, our type system
must explicitly track functoriality of types, and cocontinuity
conditions on the functors interpreting them must be appropriately
threaded throughout the model construction. We also prove that our
model satisfies an appropriate Abstraction Theorem, as well as that it
verifies all standard consequences of parametricity in the presence of
primitive nested types. In particular, we give several examples
illustrating how how our model can be used to derive useful free
theorems for programs over nested types. {\color{red} Add info about
  GADTs and implementation.}
\end{abstract}

@@

Can bind variables of arity greater than 0 in Nat-types? Improves
expressibility?  Worth it?}

@@

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;{\color{red}\ol{\gamma^0}},\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash H}$}
\TIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F {\color{red} [\ol{\gamma := H}]})\,\ol{G}$}  
\DisplayProof
\end{array}\]

@@

{\color{red} CHANGE TEXT FOR NEW $\mu$-RULE.}
Definition~\ref{def:wftypes} allows the formation of all of the
(closed {\color{red} NOT!}) nested types from the introduction:
\[\begin{array}{lll}
\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
\beta)\,\alpha\\ 
\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
\phi\,(\beta \times \beta))\,\alpha\\
\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
\times \phi\,(\phi\beta))\,\alpha 
\end{array}\]
{\color{red} The representation of $\mathit{List}\,\alpha$ as the ADT
  $\mu \beta. \onet + \alpha \times \beta$ cannot be functorial in
  $\alpha$. By contrast, the representation of $\mathit{List}\,\alpha$
  as the nested type $(\mu \phi. \lambda \beta.\,\onet + \beta \times
  \phi \beta)\,\alpha$
%Each of these types
  can be considered either functorial in $\alpha$ or not, according to
  whether $\alpha$ is in the functorial or non-functorial context in
  which this type is formed.} In particular, if $\emptyset;\alpha
\vdash \mathit{List} \,\alpha$, then $\vdash \Nat^\alpha \onet\,
(\mathit{List}\,\alpha)$ is well-formed; if $\alpha;\emptyset \vdash
\List \,\alpha$, then it is not.  Similarly, if
$\mathit{Tree}\,\alpha\,\gamma = \mu \,\beta.\, \alpha + \beta \times
\gamma \times \beta$, then $\gamma;\emptyset \vdash \Nat^\alpha
(\mathit{List}\, \alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is
well-formed, but $\emptyset;\gamma \vdash \Nat^\alpha (\mathit{List}\,
\alpha) \, (\mathit{Tree} \, \alpha\, \gamma)$ is not. {\color{red}
  KEEP? Although types can be functorial in variables of arity greater
  than $0$, the body $F$ of a type $(\mu \phi.\lambda \ol\alpha.F)\ol
  G$ cannot.} For example, the type $\mathit{GRose}\, \phi \, \alpha =
\mu\beta . \alpha \times \phi \beta$ {\color{red} can be functorial
  --- NOT!} or not in $\alpha$, but cannot be functorial in
$\phi$. Moreover, $\mathit{GRose}\,\phi\,\alpha$ cannot be the
(co)domain of a $\Nat$-type representing a natural transformation
functorial in $\phi$, and the type $\mu \phi. \lambda
\alpha. \mathit{GRose}\,\phi\,\alpha$ is not well-formed in our
system. These restrictions ensure that the set and relational
interpretations of types in Section~\ref{sec:type-interp} are
functors. In Section~\ref{sec:conclusion} we discuss how more
expressive types might be accommodated.

Definition~\ref{def:wftypes} allows well-formed types to be functorial
in no variables. Functorial variables
%in a well-formed type $\tau$
can also be demoted to non-functorial status: if\,$F[\phi :== \psi]$
is the textual replacement of $\phi$ in $F$, then $\Gamma, \psi^k;
\Phi \vdash F[\phi^k :== \psi^k]$ is derivable whenever $\Gamma; \Phi,
\phi^k \vdash F$ is. The proof is by induction on the structure of
$F$. In addition to textual replacement, we also have a
substitution operation on types.

@@

{\color{red} Check no change needed for new $\mu$-rule in next
  def. Just note that in application case the args to $\phi$ go in for
  the $\alpha$s in $F$s replacing $\phi$.}
\begin{defi}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
\vdash F$ with $|\ol\alpha| = k$, then the operation $H[\phi :=_{\ol
    \alpha} F]$ of {\em second-order type substitution along
  $\ol\alpha$} is defined by induction on $H$ as follows:
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, (G[\phi :=_{\ol{\alpha}} F]) \,(K[\phi
  :=_{\ol{\alpha}} F])\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G[\phi :=_{\ol{\alpha}}
  F])\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{defi}
\noindent
It is not hard to see that $\Gamma;\Phi \vdash H[\phi :=_{\ol{\alpha}}
  F]$. In addition, if \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma;
\ol\psi,\ol{\alpha} \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap
\ol\psi = \emptyset$, then $\Gamma,\ol\psi';\Phi \vdash H[\phi
  :=_{\ol{\alpha}} F[\ol{\psi :== \psi'}]]$.

@@

       \[\begin{array}{c}
       \AXC{$\Gamma; \ol{\alpha} \vdash F$}
       \AXC{$\Gamma; \ol{\alpha} \vdash G$}
       \AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
       \TIC{$\Gamma; \emptyset
         \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
       \DisplayProof\vspace*{-0.05in}
       \\\\
       \AXC{$\ol{\Gamma;\Phi \vdash K}$}
       \AXC{$\Gamma; \emptyset
         \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
       \TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
         G[\overline{\alpha := K}]$}
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash
           G}$}
       \TIC{$\Gamma; \emptyset
         ~|~\emptyset
         \vdash \map^{\ol{F},\ol{G}}_H :
         \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
         (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
             :=_{\ol{\beta}} G}])$} 
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi, \ol{\alpha}, {\color{red}\ol{\gamma}} \vdash H$}
       \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
         \Nat^{\ol{\beta},{\color{red}\ol{\gamma}}} H[\phi :=_{\ol{\beta}} (\mu
           \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
         \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi,\ol{\alpha}, {\color{red} \ol{\gamma}} \vdash H$}
       \AXC{$\Gamma; \ol{\beta}, {\color{red} \ol{\gamma}} \vdash F$}
       \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
         \Nat^\emptyset\; (\Nat^{\ol{\beta}, {\color{red} \ol{\gamma}}}\,H[\phi
           :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\ol{\beta},
          {\color{red} \ol{\gamma}}}\,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \end{array}\]

@@

       \caption{Well-formed terms {\color{red} Remove red $\gamma$s?
           Leave rule for map as
           is???}}\label{fig:terms} \vspace*{-0.00in}
\end{adjustbox}
%       \vspace*{-0.25in}
\end{figure*}

@@

{\color{red} Change the rest of the section to agree with new term
  formation rules.}  The ``extra'' functorial variables $\ol{\gamma}$
in the rules for $\map^{\ol{F},\ol{G}}_H$ {\color{red} , $\tin_H$, and
  $\fold^F_H$} (i.e., those variables not affected by the substitution
of $\phi$) deserve comment. They allow us to map {\color{red} or fold}
polymorphic functions over nested types. Suppose, for example, we want
to map the polymorphic function $\mathit{flatten} : \Nat^\beta
(\mathit{PTree}\,\beta)\,(\mathit{List}\,\beta)$ over a list.  Even in
the absence of extra variables the instance of $\map$ required to map
each non-functorial monomorphic instantiation of $\mathit{flatten}$
over a list of perfect trees is well-typed:
\[\begin{array}{l}
\AXC{$\Gamma;\alpha \vdash \mathit{List} \, \alpha$}
\AXC{$\Gamma;\emptyset \vdash \mathit{PTree}\,F$}
\AXC{$\Gamma;\emptyset \vdash \mathit{List}\,G$}
\TIC{$\Gamma;\emptyset \,|\, \emptyset \vdash \map^{\mathit{PTree}\,F,
    \,\mathit{List}\,G}_{\mathit{List}\,\alpha} :
  \Nat^\emptyset (\Nat^\emptyset\,(\mathit{PTree}\, F)\,(\mathit{List}\, G))\,
  (\Nat^\emptyset\,(\mathit{List}\, (\mathit{PTree}\,
  F))\,(\mathit{List}\, (\mathit{List}\, G)))$} \DisplayProof
\end{array}\]
But in the absence of $\ol \gamma$, the instance
\[\Gamma;\emptyset~|~\emptyset \vdash
\map^{\mathit{PTree}\,\beta,\mathit{List}\,\beta}_{\mathit{List}\,\alpha}
: \Nat^\emptyset ( \Nat^\beta(\mathit{PTree}\,
\beta)\,(\mathit{List}\, \beta))\, (\Nat^\beta\,(\mathit{List}\,
(\mathit{PTree}\, \beta))\,(\mathit{List}\, (\mathit{List}\,
\beta)))\] required to map the {\em polymorphic} $\mathit{flatten}$
function over a list of perfect trees is not: indeed, the functorial
contexts for $F$ and $G$ in the rule for $\map^{F,G}_H$ would have to
be empty, but because the polymorphic $\mathit{flatten}$ function is
natural in some variable, say $\delta$, it cannot possibly have a type
of the form $\Nat^\emptyset F\, G$ as would be required for it to be
the function input to $\map$. Untypeability of this instance of $\map$
is unsatisfactory in a polymorphic calculus, where we naturally expect
to be able to manipulate entire polymorphic functions rather than just
their monomorphic instances, but the ``extra'' variables $\ol \gamma$
remedy the situation, ensuring that the instance of $\map$ needed to
map the polymorphic $\mathit{flatten}$ function is typeable as
follows:

@@

{\color{red} The next discussion should still be OK.}

@@

Unfortunately, our calculus cannot express types of recursive
functions --- such as a concatenation function for perfect trees or a
zip function for bushes --- that take as inputs a nested type and an
argument of another type, both of which are parameterized over the
same variable. The fundamental issue is that recursion is expressible
only via $\fold$, which produces natural transformations in some
variables $\ol\alpha$ from $\mu$-types to other functors $F$. The
restrictions on $\Nat$-types entail that $F$ cannot itself be a
$\Nat$-type containing $\ol{\alpha}$, so, e.g., $\Nat^\alpha
(\mathit{PTree}\,\alpha) \,(\Nat^\emptyset (\mathit{PTree}\,\alpha)\,
(\mathit{PTree}\,(\alpha \times \alpha)))$ is not well-typed.
Uncurrying gives $\Nat^\alpha (\mathit{PTree}\,\alpha \times
\mathit{PTree}\,\alpha)\, (\mathit{PTree}\,(\alpha \times \alpha))$,
which is well-typed, but $\mathsf{fold}$ cannot produce a term of this
type because $\mathit{PTree}\,\alpha \times \mathit{PTree}\,\alpha$ is
not a $\mu$-type.  Our calculus can, however, express types of
recursive functions that take multiple nested types as arguments,
provided they are parameterized over disjoint sets of type variables
and the return type of the function is parameterized over only the
variables occurring in the type of its final argument.  {\color{red}
  CHECK THE FOLLOWING! SHOULD BE OK.}  Even for ADTs there is a
difference between which folds over them we can type when they are
viewed as ADTs {\color{red} (i.e., as fixpoints of first-order
  functors) EARLIER?}  versus as proper nested types {\color{red}
  (i.e., as fixpoints of higher-order functors) EARLIER?}. This is
because, in the return type of $\mathsf{fold}$, the arguments of the
$\mu$-type must be variables bound by $\Nat$.  For ADTs, the
$\mu$-type takes no arguments, making it possible to write recursive
functions, such as a concatenation function for lists of type $\alpha
; \emptyset \vdash \Nat^\emptyset \ (\mu \beta. \onet + \alpha \times
\beta)\, (\Nat^\emptyset (\mu \beta. \onet + \alpha \times \beta) \,
(\mu \beta. \onet + \alpha \times \beta))$.  This is not possible for
nested types --- even when they are semantically equivalent to ADTs.

@@

\item If \,$\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu \phi.\lambda

@@

  \setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu  \phi.\lambda
    \ol{\alpha}. H)\ol{G}} f &: &\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} 

@@

  \setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash (\mu

@@

  T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi,{\color{red}\ol\gamma} \vdash G}\rho} \to (\mu
